Structuring College Access: The Market Segment Model and College Board Geomarkets

### 2025-03-12 text/front_end_v3.qmd
intro

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased. First, the order filtered on 2020 high school graduating class. Second, under the heading "College Board Exams," the order filtered on PSAT score range of 1070 - 1180. Third, under the heading "Geography", the order filtered on 67 "Geomarkets." Geomarkets carve states and metroplitan areas into smaller geographic units and have names like, IL08 - Northwest Suburbs" and "IL09 - North Shore."

On the same day, the university also purchased orders $\#448427$ ($57,537$ names) and $\#448440$ ($78,000$ names), which filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets as order $\#448922$. On closer inspection, the orders filtered on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White than surrounding Geomarkets. In the DC/Maryland/Virginia metro area, the orders included the "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 [check] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.

@@

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting (p.9)\ldots We have begun with this celebration of sotrytelling\ldots [because] we believe that the intuitios of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research this is based opn listening carefully to what admissions officers have to say (pp. 9-10).

@@

The following section provides background about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. [We find that\ldots ?PARAGRAPH ON RESULTS?] Finally, we discuss implications for scholarship and for policy. The sociology of education should move beyond the fixation on schools and policies as the primary sorting machines. @RN527[p. 148] remind us that the organizational field includes the totality of relevant actors." Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product [CITE]. By investigating high-leverage actors and products in the organizational field, sociology can develop important insights that shape future federal and state regulatory policy.

@@

<!-- la la -->
Enrollment management is simultaneously a profession, an administrative structure, and an industry. As a profession, enrollment management (EM) integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  -->

@@

<!--

@@

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."
-->

@@

**Correlation**. The interdisciplinary literature on quantification [@RN5008] includes important contributions from the field of critical data studies [e.g., @RN4772; @RN4849]. In particular, the discussions of correlation (chapter 1) and homophily (chapter 2) by @RN4975 introduce ideas salient to the analysis the Market Segment Model and Geomarkets. Correlation measures the extent to which two or more variables move together. Predictive analytics are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"^[Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?']

@@

**Homophily**. Homophily is the idea actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together" [@mcpherson2001birds]. Homophily is a core concept of network science (writ large), in which actors (nodes) are connected to one another directly and indirectly via network ties (edges). As discussed below, @RN4982[chapter 4] *The company We Keep: Colleges and Their Competition* conducts a network analysis based on 1980 SAT score sending behavior whereby two colleges are defined as competitors if a large number of students send SAT scores to both colleges. Because network science models often draws from rational choice theory, @RN4975 argues, it assumes that homophily is the result of voluntary action by individuals.

@@

[^3]: Thus, @RN4975 [p. 82] writes, "echo chambers are not unfortunate errors, but deliberate goals" because "homophily is used to create agitated clusters of individuals whose angry similarity and overwhelming attrication to their common object of hatred both repel them from one another and glue them together."

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^4] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there.
<!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

[^4]: For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether.

@@

**Creating Geomarkets and the Market Segment Model.** In 1978, University of Pennsylvania professor Robert Zemsky, was tasked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s. <!--The geographic concepts inscribed in the Market Segment Model were based on conversations with college admissions officers. -->  The "initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools" [@RN4982,p. 4]. Quoting an admissions officer, @RN4982 [p. 11] write, "'There are only three kinds of college-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, [@RN4982] created three geographic [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. They "divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" (p. 11). These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982[pp. 11-12] briefly describes the creation of Geomarket borders:

@@

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than $35,000, percent with both parents having a BA, etc.^[For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.^[Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.] Taken together, the Market Segment Profile shows a colleges which Geomarkets possess attractrive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

**Correlations**. @RN4982[Chapter 3], *A Sense of Place: Students, Families, and Communities*, identifies the student characteristics correlated with being in the local, in-state, regional, or national market segment. The analyses identify four variables -- educational aspirations, parental education, scholastic aptitude, and family income -- that predict score sending behavior, both individually and in combination. These four variables "reflect the basic social patterns of the nation. It would have been surprising if these were not the four social variables that best explained the patterns of college choice" [@RN4982, p. 33].<!--
and "The information derived from the Market Segment Model, however, is a remarkably ordered set of data, consistent in its relationships, that reflects the basic social and economic patterning associated with the structure of college choice" (pp. 34-35). \[QUOTES REDUNDANT\]
-->

The thesis of Chapter 3 is that student demand for higher education is a function of social origin. For example, @RN4982[p. 33] state that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982 [p. 33] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time."[^6] @RN4982[p. 42] conclude that "our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations." Geomarkets differ in the relative abundance of students with particular socioeconomic characteristics, which has practical implications for recruiting. Therefore, @RN4982[p. 44] recommend that colleges target Geomarkets with desirable compositions of socioeconomic characteristics in order to reach students from desired student market segments:

> On occasion, senior spokespersons for the profession worry that students outside the main [Geo]market areas remain forgotten and hence, unchallenged. Inevitably, the increasing competition for students, the expense of travel and mailings, and internal political constraints compel institutions to concentrate their efforts where they will do the most good. The result is a natural reinforcing of the basic socioeconomic patterns that gave shape in the first place to the structure of college choice.

@@

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one anothyer.[^8] Based on these analyses, @RN4982 [p. 46] state that homophily describes the nature of competition between colleges: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe these patterns as a natural process of homophily in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

@@

[^8]: This is a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). The authors turn this into a one-mode college network that defines two institutions as being in competition with one another -- the network tie -- if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa. Next, the authors develop "tinker toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market. For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."

@@

Whereas @RN4982 identified 143 Geomarkets covering the New England, Middle States, and South region, EPS created Geomarkets for the remaining U.S. states, with 304 Geomarkets in total. @RN4988 shows the contemporary Geomarkets. Documentation and promotional material suggests that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.[^5] However, Geomarkets for New England, Middle States, and the South are identical to those developed by @RN4982.

[^5]: @RN5016 states that "geomarkets are areas within a state that represent a further segmentation of a population. Students from California don't all share the same college-going behaviors. We have accounted for this variance by segmenting the 50 states into 304 geomarkets to provide further insight into student behaviors within particular areas of individual states."

@@

EPS software may discipline colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.^[Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.

@@

Finally, we suspect that Geomarkets affect how admissions offices organize the recruiting territories assigned to each admissions officer. Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it was common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket. In those states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago [ADD TO APPENDIX], of Lake Forest College,  of Saint Mary's College [ADD TO APPENDIX], [ADD TO APPENDIX]. <!-- uc san diego appears to use geomarkets to organize high school visits: inspect the html in this page https://beatriton.ucsd.edu/register/requestschoolvisit -->

@@

<!-- 
Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:
-->

@@

We received both (a) the order summary data and (b) the de-identified prospect-level student list data for 414 orders associated with 2,549,085 prospects. We draw from these data to answer RQ2. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase.[^17]. Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

[^17]: ADD CAVEAT THAT PURCHASES EXCLUDE PROSPECT PROFILES THAT WERE PREVIOUSLY PURCHASED

>>>>

On January 9, 2019 a public research university purchased College Board *Student Search Service* order no. $448922$, thereby obtaining the contact information of out-of-state prospective students, who would subsequently receive brochures, postcards, email, and targeted social media. This was a large student list purchase, yielding $122,426$ "names." At $\$0.45$ per name, the order cost $\$55,091.70$. The order specified three "search filters" to control which names were purchased: first, the 2020 high school graduating class; second, under the heading "College Board Exams," PSAT score range of 1070 - 1180; and third, under the heading "Geography", the order filtered on 67 "Geomarkets." Geomarkets carve states and metroplitan areas into smaller geographic units and have names like, IL08 - Northwest Suburbs" and "IL09 - North Shore." On the same day, the university purchased two additional orders that were identical, except they filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$. <!--On the same day, the university also purchased orders $\#448427$ ($57,537$ names) and $\#448440$ ($78,000$ names), which filtered on PSAT score ranges of $1190 - 1260$ and $1270 - 1520$, respectively, but selected the same HS graduating class and Geomarkets as order $\#448922$.   -->

On closer inspection, the three orders selected on all Geomarkets in the Chicago-land area except for "IL10 - City of Chicago", which is significantly poorer and less White. In the DC/Maryland/Virginia metro area, the orders selected the affluent "MD02 - Montgomery Metropolitan," "MD03 - Central Maryland (w/o Baltimore)," "VA01 - Arlington & Alexandria," and "VA02 - Fairfax County" Geomarkets, but excluded the "MD05 - Prince Georges Metropolitan," "MD07 - Baltimore (Urban)," and "DC01 - District of Columbia" Geomarkets, which are poorer and have a higher share of Black residents. An analysis of 500 \[check\] student lists purchased by 9 public research universities found that 11% of purchases filtered on Geomarket [@RN5012]. Recent research suggests that student lists are surprisingly important for college access outcomes and college graduation outcomes, particularly for first-generation and underrepresented minority students [@RN4739; @RN4894]. To the extent that Geomarket borders are correlated with race and class, the Geomarket selections observed here raise concern that Geomarkets -- and other geographic search filters -- may systematically filter working class and non-white students from educational opportunities.

@@

> Admissions officers invariably are tellers of stories -- about the colleges they represent, about the colleges they attended, about each other, and about the often vagabond life of college recruiting (p.9)\ldots We have begun with this celebration of sotrytelling\ldots \[because\] we believe that the intuitios of admissions officers actually comprise a remarkably systematic body of knowledge about the college selection process\ldots Our research this is based opn listening carefully to what admissions officers have to say (pp. 9-10).

@@

The following section provides background about enrollment management and introduces salient concepts from scholarship on quantification. Second, we introduce the case, describing how @RN4982 developed Geomarkets and the Market Segment Model and how these concepts were incorporated into College Board recruiting products. Third, we describe data and methods. Fourth, we present results. [We find that\ldots ?PARAGRAPH ON RESULTS?] Finally, we discuss implications for scholarship and for policy. The sociology of education should move beyond the fixation on schools and policies as the primary sorting machines. @RN527 [p. 148] remind us that the organizational field includes the totality of relevant actors." Third-party vendors have structured college access for decades and their influence is growing. The contemporary enrollment management industry is characterized by private equity owned firms selling software-as-service platforms that utilize the same logic as @RN4982, for example EAB's "pipeline analytics" machine learning product \[CITE\]. By investigating high-leverage actors and products in the organizational field, sociology can develop important insights that shape future federal and state regulatory policy.

@@

<!-- la la -->

Enrollment management is simultaneously a profession, an administrative structure, and an industry. As a profession, enrollment management (EM) integrates techniques from marketing and economics in order to "influence the characteristics and the size of enrolled student bodies" [@RN2771, p. xiv]. As an administrative structure, the EM office typically controls the activities of admissions, financial aid, and recruiting [@RN2406]. <!--The enrollment management industry incorporates relevant actors in the organizational field, including professional associations (e.g., National Association for College Admission Counseling), and third-party servicers (e.g., College Board, EAB) that supply products and consulting solutions to colleges.  -->

@@

```{=html}
<!--

@@

[^1]: @RN3519 describes the archives of annual reports of admissions activities from a selective liberal arts college. In the 1942 report, "Section 2.5 is titled 'Secondary School Quartile Standings," a measure of where matriculants fell in the rankings of their high school classes" [@RN3519] while Section 3.11 describes the high schools visited. In the chapter entitled "Travel," @RN3519 [p. 69] describes how The College used record-keeping to inform decisions about visits to local high schools in the 2000s: "The College keeps files on hundreds of schools across the country, and around the globe, with which it had contact \ldots The files enable officers to quickly discern"how we've done here lately," \ldots with the number, relative quality, and yield rate of applicants from a particular school."
-->
```

@@

**Correlation**. The interdisciplinary literature on quantification [@RN5008] includes important contributions from the field of critical data studies [e.g., @RN4772; @RN4849]. In particular, the discussions of correlation (chapter 1) and homophily (chapter 2) by @RN4975 introduce ideas salient to the analysis the Market Segment Model and Geomarkets. Correlation measures the extent to which two or more variables move together. Predictive analytics are based on correlation and are developed in two steps. First, apply statistical techniques to previous cases (training data) in order to identify factors positively and negatively associated with an outcome of interest. Second, apply these results (e.g., regression coefficients) to future cases in order to make predictions and to assign levels of risk to each case. @RN4975 provides the example of @RN4989, who develop a method to predict sensitive personal attributes (e.g., gender, political party) based on Facebook Likes. These models predict outcomes based on correlations without requiring knowledge about underlying causal relationships. @RN4975 [p. 50] writes that "correlation grounds big data's so-called revolutionary potential. As Wired editor Chris Anderson infamously declared \ldots, big data proved that 'correlation supersedes causation, and science can advance even without coherent models, unified theories, or really any mechanistic explanation at all.'"[^1]

[^1]: Due to data limitations, predictive analytics often utilize "proxy" variables [@RN4849; @RN4975], which are variables that are correlated with a variable of interest. For example, "e-scores" utilize proxy variables to identify "people like you" and then predict your buying behavior based on the past buying behavior of people like you. @RN4849 [p. 146] states that "the modelers for e-scores have to make do with trying to answer the question 'How have people like you behaved in the past?' when ideally they would ask, 'How have you behaved in the past?'" We observe similar behavior in market research about college access, when enrollment managers use 'which colleges did students near you consider' as a proxy for, 'which colleges would you consider?'

@@

**Homophily**. Homophily is the idea actors who share common characteristics are likely to form connections with one another, or "birds of a feather flock together" [@mcpherson2001birds]. Homophily is a core concept of network science (writ large), in which actors (nodes) are connected to one another directly and indirectly via network ties (edges). As discussed below, @RN4982 [chapter 4] *The company We Keep: Colleges and Their Competition* conducts a network analysis based on 1980 SAT score sending behavior whereby two colleges are defined as competitors if a large number of students send SAT scores to both colleges. Because network science models often draws from rational choice theory, @RN4975 argues, it assumes that homophily is the result of voluntary action by individuals.

@@

Homophily is central to market research products that categorize customers. Geodemography emerged in the 1970s as a branch of market research that estimates the behavior of consumers based on where they live [@RN4878]. Market segments are subgroups within a larger market that have similar consumer demand. Early geodemographic classifications of consumers (e.g., PRIZM by Claritas Corporation) were derived from publicly available Census data, which disaggregated data to the zip code level. The Claritas Corporation had a financial incentive to argue that people living near one another share similar consumer preferences because geographic localities could then be categorized into market segments that would be useful for direct mail marketing campaigns [@RN4957]. Later, the development of individual credit scores (e.g., FICO score) enabled merchants to classify consumers into many, fine-grained groups [@RN4877]. @RN4810 introduce the concept "classification situations" to describe the expansion of actuarial techniques to categorize customers into many, ordinally ranked groups. Merchants and lenders began tying these classifications to tiered products that targets different consumer groups with different levels of benefits and costs [@RN5015].[^3] Classification situations engender markets where a vertical hierarchy of products are matched to a vertical hierarchy of consumers. We observe similar processes in our case study. @RN4982 categorizes students into four market segments -- local, in-state, regional, and national -- based on their score sending behavior and then evaluates the attractiveness of each Geomarket based on how many "regional" and "national" students live there. <!--
ORIGINAL TEXT ABOUT GEODEMOGRAPHIC CLASSIFICATION SYSTEMS

@@

[^3]: For example, "payday loans" charge high interest rates to consumer groups that were previously denied credit altogether.

@@

**Creating Geomarkets and the Market Segment Model.** In 1978, University of Pennsylvania professor Robert Zemsky, was tasked by the President to figure out, "'Who thinks about Penn?'" and "'What other institutions do they think about when they think about us?" [@RN4982, p. x]. Zemsky began working with the Market Research Committee of the Consortium on Financing Higher Education (COFHE), a consortium of 30 selective private universities founded in the mid-1970s. <!--The geographic concepts inscribed in the Market Segment Model were based on conversations with college admissions officers. --> The "initial task was to define enrollment markets in a manner consistent with admissions officers' intuitive understanding of student pools" [@RN4982,p. 4]. Quoting an admissions officer, @RN4982 [p. 11] write, "'There are only three kinds of college-bound students: those who want to live at home, those who want to live on campus but bring their laundry home, and those who want to go far enough from home that Mom and Dad can't visit without calling first.'" As such, [@RN4982] created three geographic [@RN4982, p. 11], region, state, and community. The three regions were New England, Middle States, and the South. They "divided each state into as few as two and as many as thirty community-based enrollment markets or pools, for a total of 143 separate markets" (p. 11). These enrollment markets, later called Geomarkets, were intended to be consistent with the conception of a catchment market from the perspective of admissions counselors. @RN4982 [pp. 11-12] briefly describes the creation of Geomarket borders:

@@

The two primary outputs of the Market Segment Model are the (1) Market Segment Profile and (2) the Institutional Profile. Appendix A describes these outputs in more detail. Both outputs are created separately for each Geomarket. For each market segment (local, in-state, regional, national), the Market Segment Profile shows the number of students, their average SAT score, the percent aspiring to a BA+, percent with family in come greater than \$35,000, percent with both parents having a BA, etc.[^4] The Institutional Profile shows the number of students who send scores to a particular institution -- separately by market segment -- and which majors these students are interested in. A university could obtain this for their own institution or for a competitor.[^5] Taken together, the Market Segment Profile shows a colleges which Geomarkets possess attractrive student market segments while the Institutional Profile shows the extent to which students in a particular Geomarket are interested in your college or a competitor college. These outputs became the basis for the Enrollment Planning Service (EPS) software.

[^4]: For example, Appendix @tbl-market-segment-characteristics shows how many students in "CT3 -- Fairfield County" are defined as regional or national based on their SAT score-sending behavior.

[^5]: Appendix @tbl-sample-institutional-profile reproduces a partial, simplified version of @RN4982 [Figure 2.3], the Institutional Profile of an anonymous college for students from Fairfield County, CT. This table shows that 69 Regional students and 109 National students from Fairfield County sent SAT scores to the college.

**Correlations**. @RN4982 [Chapter 3], *A Sense of Place: Students, Families, and Communities*, identifies the student characteristics correlated with being in the local, in-state, regional, or national market segment. The analyses identify four variables -- educational aspirations, parental education, scholastic aptitude, and family income -- that predict score sending behavior, both individually and in combination. These four variables "reflect the basic social patterns of the nation. It would have been surprising if these were not the four social variables that best explained the patterns of college choice" [@RN4982, p. 33].<!--
and "The information derived from the Market Segment Model, however, is a remarkably ordered set of data, consistent in its relationships, that reflects the basic social and economic patterning associated with the structure of college choice" (pp. 34-35). \[QUOTES REDUNDANT\]
-->

The thesis of Chapter 3 is that student demand for higher education is a function of social origin. For example, @RN4982 \[p. 33\] state that "these data allow us to say with considerable confidence that local and in-state students are not likely to come from families in which both parents have received college educations" and "the implication is simply that college-educated parents instill in their children more wide-ranging educational aspirations." Commenting on family income, @RN4982 [p. 33] write that "we could predict that all local students would come from moderate-income or low-income families and be wrong only 5.5 percent of the time."[^6] @RN4982 \[p. 42\] conclude that "our research has simply demonstrated what everyone has always known: communities with high levels of family income and parental education are also communities in which students have higher than average SATs and more far-reaching aspirations." Geomarkets differ in the relative abundance of students with particular socioeconomic characteristics, which has practical implications for recruiting. Therefore, @RN4982 [p. 44] recommend that colleges target Geomarkets with desirable compositions of socioeconomic characteristics in order to reach students from desired student market segments:

@@

> On occasion, senior spokespersons for the profession worry that students outside the main \[Geo\]market areas remain forgotten and hence, unchallenged. Inevitably, the increasing competition for students, the expense of travel and mailings, and internal political constraints compel institutions to concentrate their efforts where they will do the most good. The result is a natural reinforcing of the basic socioeconomic patterns that gave shape in the first place to the structure of college choice.

**Homophily**. *The Company We Keep: Colleges and Their Competition* [@RN4982, Chapter 4] conducts a network analysis to determine which institutions are in competition with one anothyer.[^7] Based on these analyses, @RN4982 \[p. 46\] state that homophily describes the nature of competition between colleges: "we draw a fundamental conclusion about the structure of college choice: collegiate competition occurs principally between like institutions." Subsequent analyses investigate the tuition price and socioeconomic composition of institutions in competition with one another. Private selective colleges and private flagship universities compete directly for students, charge the highest prices, and enroll students with the highest socioeconomic status. The authors argue that like-colleges compete for like-students as defined by socioeconomic characteristics. @RN4982 [p. 72] describe these patterns as a natural process of homophily in which a vertical socioeconomic hierarchy of students is matched to a vertical hierarchy of universities:

[^7]: This is a "two-mode" social network in which students (mode 1) send SAT scores -- the network tie -- to colleges (mode 2). The authors turn this into a one-mode college network that defines two institutions as being in competition with one another -- the network tie -- if at least 15% of students who sent SAT scores to one institution also sent scores to the other institution and vice-versa. Next, the authors develop "tinker toy" diagrams that show which institutions are connected to one another. These diagrams are drawn separately for each student segment -- local, in-state, regional, and national -- and separately for each geomarket, such that the analyses convey which institutions compete with one another for which student segments in each local market. For example, describing the Figure 4.4 "Structure of Fairfield County Regional Market," [@RN4982, p. 54] state that "competitive overlap, moreover, is often confined to institutions belonging to the same \[Carnegie\] type as well sector. For example, public flagships compete primarily with other public flagships; private standard colleges, with other private standard colleges; Catholic institutions, with other Catholic institutions."

@@

@@

Whereas @RN4982 identified 143 Geomarkets covering the New England, Middle States, and South region, EPS created Geomarkets for the remaining U.S. states, with 304 Geomarkets in total. @RN4988 shows the contemporary Geomarkets. Documentation and promotional material suggests that geomarket borders were chosen based on a combination of formal geographic borders (e.g., counties) as well as proprietary College Board data designed to identify geographic areas with different college-going behaviors.[^8] However, Geomarkets for New England, Middle States, and the South are identical to those developed by @RN4982.

[^8]: @RN5016 states that "geomarkets are areas within a state that represent a further segmentation of a population. Students from California don't all share the same college-going behaviors. We have accounted for this variance by segmenting the 50 states into 304 geomarkets to provide further insight into student behaviors within particular areas of individual states."

@@

EPS software may discipline colleges to approach recruiting in a manner consistent with the Market Segment Model. Drawing from promotional literature [e.g., @RN5019; @RN1621], @RN4980, and background conversations with enrollment managers, the practical purpose of EPS was to inform the "travel schedule" of admissions recruiters. EPS promotional material and user guides encourages users to begin by identifying which Geomarkets they will recruit from.[^9] Second, EPS users decide which high schools they will visit within selected Geomarkets. Based on the principle of homophily, the Market Segment Model suggests that selective colleges should target Geomarkets with large numbers of affluent, college educated households, while low-income communities are left to local four-year and community colleges.

[^9]: Software documentation by @RN5017 states that, "EPS market codes are proprietary market codes owned by the College Board and are used to categorize external organizations and people into geographical areas\ldots. Some admissions offices use EPS market codes to focus their recruiting efforts in geographic areas in which they believe they will be the most successful."

@@

Finally, we suspect that Geomarkets affect how admissions offices organize the recruiting territories assigned to each admissions officer. Geomarkets were created to mirror the territories of admissions officers [@RN4982]. In turn, as Geomarkets became more salient, admissions offices often structured their territories around Geomarkets. On background, admissions officers and enrollment consultants told us it was common parlance to hear an admissions officer say something like, "I recruit 'PA 2'," which refers to the "Chester County, PA Geomarket. In those states a college recruits heavily from, admissions officers are often assigned specific Geomarkets as their territory. This can be seen in the Chicago-land recruiting territories of University of Chicago \[ADD TO APPENDIX\], of Lake Forest College, of Saint Mary's College \[ADD TO APPENDIX\], \[ADD TO APPENDIX\]. <!-- uc san diego appears to use geomarkets to organize high school visits: inspect the html in this page https://beatriton.ucsd.edu/register/requestschoolvisit -->

@@

```{=html}
<!-- 
Additionally, @RN4982 created geomarkets in service of the Market Segment Model. The Market Segment Model argues that socioeconomic status drives whether students belong to the local, in-state, regional, or national market segment. In turn, geomarkets are described by the relative numbers of local, in-state, regional, or national students, which is substantially a function of socioeconomic status. The language of @RN4982 [pp. 11-12, quoted above] suggests that geomarket borders within large metropolitan areas may have been drawn in ways that separate affluent from less affluent communities. We present the following hypothesis:
-->
```

@@

We received both (a) the order summary data and (b) the de-identified prospect-level student list data for 414 orders associated with 2,549,085 prospects. We draw from these data to answer RQ2. Although these lists were purchased by individual universities, the set of prospects included in each list is a function of the search criteria specified for that student list purchase.[^18]. Therefore, we utilize these data not to analyze the behavior of individual universities, but to identify which prospective students are included when a particular set of search criteria are selected.

[^18]: ADD CAVEAT THAT PURCHASES EXCLUDE PROSPECT PROFILES THAT WERE PREVIOUSLY PURCHASED
